---
title: "ReproducibleResearchProject2"
output: html_document
---
# Effects of weather events in the United States  

##Synopsis  
  
This report aims to identify the weather events in the United States that are most harmful with respect to population health and have the greatest economic consequences.   To identify these items, data from the U.S. National Oceanic and Atmospheric Administrations's (NOAA) storm database is used.  The data tracks characteristics of major storms and weather events in the U.S. including date, location, fatalities, injuries, and property damage.  From this data we identified that between January 1996 and November 2011.....

Packages needed for Analysis
```{r}
library(readr)
```


## Data Processing  
  
The website for this course provided Storm Data from NOAA for the years 1950 to 2011.  [link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)   
  
### Downloading and Reading the Data  
```{r setoptions, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="data.csv.bz2")
data <- read_csv("data.csv.bz2")
```
  
  
## Results  

