---
title: "ReproducibleResearchProject2"
output: html_document
---
# Effects of weather events in the United States  

##Synopsis  
  
This report aims to identify the weather events in the United States that are most harmful with respect to population health and have the greatest economic consequences.   To identify these items, data from the U.S. National Oceanic and Atmospheric Administrations's (NOAA) storm database is used.  The data tracks characteristics of major storms and weather events in the U.S. including date, location, fatalities, injuries, and property damage.  From this data we identified 8 weather events causeing the greatest threat to opulation health between January 1996 and November 2011 while floods caused the greatest economic impact.

Packages needed for Analysis
```{r message=FALSE, warning=FALSE}
library(readr)
library(lubridate)
library(dplyr)
library(ggplot2)
```


## Data Processing  
  
The website for this course provided Storm Data from NOAA for the years 1950 to 2011.  [Data is available here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)   
  
### Downloading and Reading the Data  
```{r setoptions, cache=TRUE, message=FALSE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile="data.csv.bz2")
data <- read_csv("data.csv.bz2")
```
  
###Pairing down to the data needed  
Filter data to start with 1996 when all events began being recorded to prevent inflated sums for earlier timeframes when not all events were recorded.  The analysis does not require date, time, who reported the event, or remarks.  Maintaining state to as we are looking at effects across the US, but the more specific location fields are not needed.  
```{r set options, message=FALSE}
needed<-mutate(data,date=mdy_hms(BGN_DATE)) %>%
        mutate(year=year(date)) %>%
        filter(year>1995) %>%
        select(7,8, 23,24,25,26,27,28)
```

###Identify unique event types
```{r}
needed<-mutate(needed, events=tolower(EVTYPE))
needed$events[needed$events=="tstm wind"]<-"thunderstorm wind"
needed$events[needed$events=="extreme cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="extreme windchill"]<- "extreme cold/wind chill"
needed$events[needed$events=="rip currents"]<-"rip current"
needed$events[needed$events=="minor flooding"]<-"flood"
needed$events[needed$events=="storm surge"]<-"storm surge/tide"
needed$events[needed$events=="rough surf"]<-"high surf"
needed$events[needed$events=="heavy surf"]<-"high surf"
needed$events[needed$events=="snow"]<-"heavy snow"
needed$events[needed$events=="freeze"]<-"frost/freeze"
needed$events[needed$events=="snow squalls"]<-"heavy snow"
needed$events[needed$events=="coastal flooding"]<-"coastal flood"

needed$events[needed$events=="coastalflood"]<-"coastal flood"
needed$events[needed$events=="erosion/cstl flood"]<-"coastal flood"
needed$events[needed$events=="wet month"]<-"heavy rain"
needed$events[needed$events=="wet year"]<-"heavy rain"
needed$events[needed$events=="tidal flooding"]<-"coastal flood"
needed$events[needed$events=="river flooding"]<-"flood"
needed$events[needed$events=="damaging freeze"]<-"frost/freeze"
needed$events[needed$events=="hurrican"]<-"hurricane (typhoon)"
needed$events[needed$events=="beach erosion"]<-"storm surge/tide"
needed$events[needed$events=="flood/flash flood"]<-"flood"
needed$events[needed$events=="icy roads"]<-"ice storm"
needed$events[needed$events=="high surg"]<-"storm surge/tide"
needed$events[needed$events=="heavy rain/high surg"]<-"storm suge/tide"
needed$events[needed$events=="rain damage"]<-"heavy rain"
needed$events[needed$events=="ice jam"]<-"ice storm"
needed$events[needed$events=="unseasonable cold"]<-"cold/wind chill"
needed$events[needed$events=="early frost"]<-"frost/freeze"
needed$events[needed$events=="wintry mix"]<-"winter storm"
needed$events[needed$events=="heavy rain/high surf"]<-"storm surge/tide"
needed$events[needed$events=="blowing snow"]<-"heavy snow"
needed$events[needed$events=="street flooding"]<-"flood"
needed$events[needed$events=="excessive cold"]<-"cold/wind chill"
needed$events[needed$events=="torrential rainfall"]<-"heavy rain"
needed$events[needed$events=="late-season snowfall"]<-"heavy snow"
needed$events[needed$events=="hurrican edouard"]<-"hurricane (typhoon)"
needed$events[needed$events=="strong winds"]<-"strong wind"
needed$events[needed$events=="record warm temps."]<-"excessive heat"
needed$events[needed$events=="extended cold"]<-"cold/wind chill"
needed$events[needed$events=="snow accumulation"]<-"heavy snow"
needed$events[needed$events=="heavy snow shower"]<-"heavy snow"
needed$events[needed$events=="late snow"]<-"heavy snow"
needed$events[needed$events=="record may snow"]<-"heavy snow"
needed$events[needed$events=="record winter snow"]<-"heavy snow"
needed$events[needed$events=="heavy precipitation"]<-"heavy rain"
needed$events[needed$events=="late season snowfall"]<-"heavy snow"
needed$events[needed$events=="gusty wind"]<-"strong wind"
needed$events[needed$events=="small hail"]<-"hail"
needed$events[needed$events=="mixed precip"]<-"winter weather"
needed$events[needed$events=="black ice"]<-"ice storm"
needed$events[needed$events=="snow and ice"]<-"winter storm"
needed$events[needed$events=="cold"]<-"cold/wind chill"
needed$events[needed$events=="freezing spray"]<-"freezing fog"
needed$events[needed$events=="heatburst"]<-"heat"
needed$events[needed$events=="hail/wind"]<-"hail"
needed$events[needed$events=="hail(0.75)"]<-"hail"
needed$events[needed$events=="urban flooding"]<-"flood"
needed$events[needed$events=="urban flood"]<-"flood"
needed$events[needed$events=="volcanic ashfall"]<-"volcanic ash"
needed$events[needed$events=="sleet storm"]<-"sleet"
needed$events[needed$events=="heavy surf/high surg"]<-"storm surge/tide"
needed$events[needed$events=="hurrican/typhoon"]<-"hurricane (typhoon)"
needed$events[needed$events=="high surf advisories"]<-"storm surge/tide"
needed$events[needed$events=="gusty thunderstorm wind"]<-"thunderstorm wind"
needed$events[needed$events=="drowning"]<-"flood"
needed$events[needed$events=="ice on road"]<-"ice storm"
needed$events[needed$events=="hazardous surg"]<-"storm surge/tide"
needed$events[needed$events=="winter weather/mix"]<-"winter weather"
needed$events[needed$events=="wind chill"]<-"cold/wind chill"
needed$events[needed$events=="smoke"]<-"dense smoke"
needed$events[needed$events=="non-tstm wind"]<-"strong wind"
needed$events[needed$events=="accumulated snowfall"]<-"heavy snow"
needed$events[needed$events=="non tstm wind"]<-"strong wind"
needed$events[needed$events=="blowing dust"]<-"dust storm"
needed$events[needed$events=="excessive heat/drought"]<-"drought"
needed$events[needed$events=="tornado debris"]<-"tornado"
needed$events[needed$events=="dust devel"]<-"dust devil"
needed$events[needed$events=="very warm"]<-"excessive heat"
needed$events[needed$events=="thunderstorm wind (g40)"]<-"thunderstorm wind"
needed$events[needed$events=="locally heavy rain"]<-"heavy rain"
needed$events[needed$events=="wind gusts"]<-"strong wind"
needed$events[needed$events=="abnormally dry"]<-"drought"
needed$events[needed$events=="patchy ice"]<-"ice storm"
needed$events[needed$events=="gusty thunderstorm winds"]<-"thunderstorm wind"
needed$events[needed$events=="astronomical high tide"]<-"storm surge/tide"
needed$events[needed$events=="cstl flooding/erosion"]<-"coastal flood"
needed$events[needed$events=="red flag criteria"]<-"storm surge/tide"
needed$events[needed$events=="rogue wave"]<-"high surf"
needed$events[needed$events=="brush fire"]<-"wildfire"
needed$events[needed$events=="hazardous surg"]<-"storm surge/tide"
needed$events[needed$events=="wnd"]<-"high wind"
needed$events[needed$events=="very dry"]<-"drought"
needed$events[needed$events=="heavy rain effects"]<-"heavy rain"
needed$events[needed$events=="cold weather"]<-"cold/wind chill"
needed$events[needed$events=="hurricane/typhoon"]<-"hurricane (typhoon)"
needed$events[needed$events=="tstm wind g45"]<-"thunderstorm wind"
needed$events[needed$events=="non severe hail"]<-"hail"
needed$events[needed$events=="warm weather"]<-"heat"
needed$events[needed$events=="snow advisory"]<-"heavy snow"
needed$events[needed$events=="snow/freezing rain"]<-"winter weather"
needed$events[needed$events=="marine tstm wind"]<-"marine thunderstorm wind"
needed$events[needed$events=="ice roads"]<-"ice storm"
needed$events[needed$events=="dry"]<-"drought"
needed$events[needed$events=="unusually cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="snowmelt flooding"]<-"flood"
needed$events[needed$events=="hazardous surf"]<-"high surf"
needed$events[needed$events=="winter weather mix"]<-"winter weather"
needed$events[needed$events=="heavy surf/high surf"]<-"high surf"
needed$events[needed$events=="hurrican (typhoon)"]<-"hurricane (typhoon)"
needed$events[needed$events=="snow showers"]<-"winter weather"
needed$events[needed$events=="high water"]<-"flood"
needed$events[needed$events=="driest month"]<-"drought"
needed$events[needed$events=="record cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="late season hail"]<-"hail"
needed$events[needed$events=="dryness"]<-"drought"
needed$events[needed$events=="winter mix"]<-"winter weather"
needed$events[needed$events=="flood/flash/flood"]<-"flash flood"
needed$events[needed$events=="record precipitation"]<-"heavy rain"
needed$events[needed$events=="unseasonal cold temp"]<-"extreme cold/wind chill"
needed$events[needed$events=="non-severe wind damage"]<-"strong wind"
needed$events[needed$events=="prolong warmth"]<-"excessive heat"
needed$events[needed$events=="record low rainfall"]<-"drought"
needed$events[needed$events=="falling snow/ice"]<-"winter weather"
needed$events[needed$events=="monthly precipitation"]<-"heavy rain"
needed$events[needed$events=="extreme windchill temperatures"]<-"extreme cold/wind chill"
needed$events[needed$events=="extreme wind chill"]<-"extreme cold/wind chill"
needed$events[needed$events=="heavy rainfall"]<-"heavy rain"
needed$events[needed$events=="excessive snow"]<-"heavy snow"
needed$events[needed$events=="dry conditions"]<-"drought"
needed$events[needed$events=="remnants of floyd"]<-"hurricane (typhoon)"
needed$events[needed$events=="record cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="high seas"]<-"high surf"
needed$events[needed$events=="rough seas"]<-"high surf"
needed$events[needed$events=="unusually warm"]<-"excessive heat"
needed$events[needed$events=="extreme windchill temperatures"]<-"extreme cold/wind chill"
needed$events[needed$events=="light freezing rain"]<-"winter weather"
needed$events[needed$events=="record snow"]<-"heavy snow"
needed$events[needed$events=="high surf advisory"]<-"high surf"
needed$events[needed$events=="unseasonal low temp"]<-"extreme cold/wind chill"
needed$events[needed$events=="record rainfall"]<-"heavy rain"
needed$events[needed$events=="hot weather"]<-"heat"
needed$events[needed$events=="volcanic eruption"]<-"volcanic ash"
needed$events[needed$events=="first frost"]<-"frost/freeze"
needed$events[needed$events=="excessively dry"]<-"drought"
needed$events[needed$events=="heavy seas"]<-"high surf"
needed$events[needed$events=="flash flood/flood"]<-"flash flood"
needed$events[needed$events=="flash flooding"]<-"flash flood"
needed$events[needed$events=="record dryness"]<-"drought"
needed$events[needed$events=="wind and wave"]<-"high surf"
needed$events[needed$events=="record cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="severe thunderstorm"]<-"heavy rain"
needed$events[needed$events=="excessive rainfall"]<-"heavy rain"
needed$events[needed$events=="coastal flooding/erosion"]<-"coastal flood"
needed$events[needed$events=="hyperthermia/exposure"]<-"excessive heat"
needed$events[needed$events=="ice pellets"]<-"ice storm"
needed$events[needed$events=="record cool"]<-"extreme cold/wind chill"
needed$events[needed$events=="patchy dense fog"]<-"dense fog"
needed$events[needed$events=="record cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="cool spell"]<-"cold/wind chill"
needed$events[needed$events=="snow/sleet"]<-"winter storm"
needed$events[needed$events=="urban/street flooding"]<-"flood"
needed$events[needed$events=="coastal erosion"]<-"coastal flood"
needed$events[needed$events=="unusual/record warmth"]<-"excessive heat"
needed$events[needed$events=="bitter wind chill temperatures"]<-"extreme cold/wind chill"
needed$events[needed$events=="bitter wind chill"]<-"extreme cold/wind chill"
needed$events[needed$events=="record warm"]<-"heat"
needed$events[needed$events=="cold wind chill temperatures"]<-"cold/wind chill"
needed$events[needed$events=="cold temperatures"]<-"cold/wind chill"
needed$events[needed$events=="coastal  flooding/erosion"]<-"coastal flood"
needed$events[needed$events=="unseasonably warm year"]<-"excessive heat"
needed$events[needed$events=="light snow/freezing precip"]<-"winter weather"
needed$events[needed$events=="record  cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="high wind (g40)"]<-"high wind"
needed$events[needed$events=="tstm wind (g45)"]<-"thunderstorm wind"
needed$events[needed$events=="tstm wind (g35)"]<-"thunderstorm wind"
needed$events[needed$events=="dry weather"]<-"drought"
needed$events[needed$events=="abnormal warmth"]<-"excessive heat"
needed$events[needed$events=="unusual warmth"]<-"excessive heat"
needed$events[needed$events=="tstm"]<-"thunderstorm wind"
needed$events[needed$events=="wintery mix"]<-"winter weather"
needed$events[needed$events=="hot spell"]<-"excessive heat"
needed$events[needed$events=="unseasonably hot"]<-"excessive heat"
needed$events[needed$events=="tstm wind  (g45)"]<-"thunderstorm wind"
needed$events[needed$events=="tstm winds"]<-"thunderstorm wind"
needed$events[needed$events=="high swells"]<-"storm surge/tide"
needed$events[needed$events=="high winds"]<-"high wind"
needed$events[needed$events=="dry spell"]<-"drought"
needed$events[needed$events=="unseasonal rain"]<-"heavy rain"
needed$events[needed$events=="prolonged rain"]<-"heavy rain"
needed$events[needed$events=="early rain"]<-"heavy rain"
needed$events[needed$events=="heavy surf/wind"]<-"high surf"
needed$events[needed$events=="high  swells"]<-"high surf"
needed$events[needed$events=="typhoon"]<-"hurricane (typhoon)"
needed$events[needed$events=="hypothermia/exposure"]<-"extreme cold/wind chill"
needed$events[needed$events=="record high"]<-"excessive heat"
needed$events[needed$events=="coastal storm"]<-"coastal flood"
needed$events[needed$events=="funnel clouds"]<-"funnel cloud"
needed$events[needed$events=="blizzard summary"]<-"blizzard"
needed$events[needed$events=="ice"]<-"ice storm"
needed$events[needed$events=="icestorm/blizzard"]<-"blizzard"
needed$events[needed$events=="tstm wind and lightning"]<-"thunderstorm wind"
needed$events[needed$events=="excessive rain"]<-"heavy rain"
needed$events[needed$events=="heavy surf and wind"]<-"high surf"
needed$events[needed$events=="mild and dry pattern"]<-"drought"
needed$events[needed$events=="flood/strong wind"]<-"flood"
needed$events[needed$events=="prolong cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="cold and frost"]<-"frost/freeze"
needed$events[needed$events=="agricultural freeze"]<-"frost/freeze"
needed$events[needed$events=="urban/sml stream fldg"]<-"flood"
needed$events[needed$events=="late freeze"]<-"frost/freeze"
needed$events[needed$events=="coastalstorm"]<-"coastal flood"
needed$events[needed$events=="waterspouts"]<-"hurricane (typhoon)"
needed$events[needed$events=="blow-out tide"]<-"storm surge/tide"
needed$events[needed$events=="snow and sleet"]<-"winter storm"
needed$events[needed$events=="tstm wind (g40)"]<-"thunderstorm wind"
needed$events[needed$events=="tstm wnd"]<-"thunderstorm wind"
needed$events[needed$events=="lake effect snow"]<-"lake-effect snow"
needed$events[needed$events=="record snowfall"]<-"heavy snow"
needed$events[needed$events=="urban/small strm fldg"]<-"flood"
needed$events[needed$events=="freezing rain/sleet"]<-"sleet"
needed$events[needed$events=="frost"]<-"frost/freeze"
needed$events[needed$events=="rain (heavy)"]<-"heavy rain"
needed$events[needed$events=="mixed precipitation"]<-"winter weather"
needed$events[needed$events=="sleet/freezing rain"]<-"sleet"
needed$events[needed$events=="blow-out tides"]<-"storm surge/tide"
needed$events[needed$events=="tstm wind 40"]<-"thunderstorm wind"
needed$events[needed$events=="hard freeze"]<-"frost/freeze"
needed$events[needed$events=="river flood"]<-"flood"
needed$events[needed$events=="tstm wind 45"]<-"thunderstorm wind"
needed$events[needed$events=="tstm wind (41)"]<-"thunderstorm wind"
needed$events[needed$events=="record warmth"]<-"excessive heat"
needed$events[needed$events=="unseasonable warm"]<-"excessive heat"
needed$events[needed$events=="unseasonable cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="tstm heavy rain"]<-"heavy rain"
needed$events[needed$events=="snow/ice"]<-"winter weather"
needed$events[needed$events=="volcanic ash plume"]<-"volcanic ash"
needed$events[needed$events=="thundersnow shower"]<-"heavy snow"
needed$events[needed$events=="unseasonably warm"]<-"excessive heat"
needed$events[needed$events=="unseasonably dry"]<-"drought"
needed$events[needed$events=="unseasonable cold"]<-"extreme cold/wind chill"
needed$events[needed$events=="freezing drizzle"]<-"sleet"
needed$events[needed$events=="record heat"]<-"excessive heat"
needed$events[needed$events=="sml stream fld"]<-"flood"
needed$events[needed$events=="cold temperature"]<-"cold/wind chill"
needed$events[needed$events=="heat wave"]<-"heat"
needed$events[needed$events=="saharan dust"]<-"dust storm"
needed$events[needed$events=="rain"]<-"heavy rain"
needed$events[needed$events=="heavy snow squalls"]<-"heavy snow"
needed$events[needed$events=="early snowfall"]<-"heavy snow"
needed$events[needed$events=="unseasonably cool"]<-"extreme cold/wind chill"
needed$events[needed$events=="cold and snow"]<-"winter storm"
needed$events[needed$events=="unseasonably cold"]<-"extreme cold/wind chill"

needed$events[needed$events=="wind"]<-"high wind"
needed$events[needed$events=="hot and dry"]<-"drought"
needed$events[needed$events=="hurricane edouard"]<-"hurricane (typhoon)"
needed$events[needed$events=="ice/snow"]<-"winter storm"
```

###After adjusting the event type for many, I filtered out rows that could not easily by mapped to one of the 48 unique event types in the supplemental material [located here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
```{r}
small<-filter(needed, events=="astronomical low tide" | events=="avalanche"|events=="blizzard"|events=="coastal flood"|events=="cold/wind chill"| events=="debris flow"|events=="dense fog"| events=="dense smoke"| events=="drought"|events=="dust devil"|events=="dust storm"| events=="excessive heat"|events=="extreme cold/wind chill"|events=="flash flood"| events=="flood"| events=="frost/freeze"| events=="funnel cloud"| events=="freezing fog"|events=="hail"| events=="heat"| events=="heavy rain"|events=="heavy snow"|events=="high surf"|events=="high wind"|events=="hurricane (typhoon)"|events=="ice storm"|events=="lake-effect snow"|events=="lakeshore flood"|events=="lightning"|events=="marine hail"|events=="marine high wind"|events=="marine strong wind"|events=="marine thunderstorm wind"|events=="rip current"|events=="seiche"|events=="storm surge/tide"|events=="strong wind"| events=="thunderstorm wind"| events=="tornado"|events=="tropical depression"| events=="tsunami"|events=="volcanic ash"|events=="waterspout"| events=="wildfire"| events=="winter storm"| events =="winter weather")
```

  
## Results  
###Population Health

In order to determine how harmful weather events are to ppulation health, we need to look at the fatalaties and injuries that occur. First, I aggregate the events by the sum of fatalities and injuries.  
```{r}
harm<-aggregate(small[3:4], by=small['events'], sum)


```
Next, I remove events that have no fatalities or injuries.   
```{r}
harm<-harm%>%
        filter(FATALITIES>0 & INJURIES>0)

```
Next, I look at a summary of harm and decide to filter by the mean of each resulting in 8 weather events causing the greatest harm to population health.
```{r}
harmsummary<-summary(harm)
harm%>%
        filter(FATALITIES>187 & INJURIES>1240)
```

###Economic Consequences  
In order to determine the greatest economic consequence of weather events we first need to compute the dollar value of the crop damage and property damage.  

First, we can remove any events that have no damage.
```{r}
econ<- small%>%
        filter(CROPDMG>0 & PROPDMG>0)
```

Next, we need adjust the dollar values of the damage to be comparable.  
```{r}

propk<- econ%>%
        filter(PROPDMGEXP=="K") %>%
        mutate(prop=PROPDMG*1000)  

propm<-econ%>%
        filter(PROPDMGEXP=="M")%>%
        mutate(prop=PROPDMG*1000000)

propB<-econ%>%
        filter(PROPDMGEXP=="B") %>%
        mutate(prop=PROPDMG*1000000000)

cropk<- econ%>%
        filter(CROPDMGEXP=="K") %>%
        mutate(prop=CROPDMG*1000)  

cropm<-econ%>%
        filter(CROPDMGEXP=="M")%>%
        mutate(prop=CROPDMG*1000000)

cropB<-econ%>%
        filter(CROPDMGEXP=="B") %>%
        mutate(prop=CROPDMG*1000000000)
```

Next, we combine those to get an aggregate of property damage per event type.
```{r}
prop<-rbind(propk, propm, propB)
prop<-aggregate(prop$prop, by=prop['events'], sum)
```

Then, we combine to get an aggregate of crop damage per event type
```{r}
crop<-rbind(cropk, cropm, cropB)
crop<-aggregate(crop$prop, by=crop['events'], sum)
```

Combine the two into one object.
```{r}
damage<-full_join(prop, crop, by="events")
```

Combine the damage values.
```{r}
damage<- damage%>%
        mutate(cost=x.x+x.y)%>%
        arrange(desc(cost))
```

Plot the results
```{r}
par(mar=c(5,12,4,2))
barplot(log(damage$cost), horiz = TRUE, main="Economic Damage from US Weather Events", sub="Floods have caused the most economic damage.", names.arg = damage$events, las=2, col=grey.colors(22),axes=FALSE)
```

